---
title: "Assignment 2: Geometrically Weighted Terms in ERGMs"
author: "-"
format: pdf
---

# Task 1: Geometrically Weighted Terms in ERGMs
## (1) Change statistic of gwodegree

The **gwodegree** effect is defined as:
$$s_{GWD}(x, \alpha) = e^\alpha \sum_{k=1}^{n} \{1 - (1 - e^{-\alpha})^k\} D_k^{out}(x)$$
where $D_k^{out}(x)$ is the number of nodes with exactly out-degree $k$ in network $x$, and $\alpha$ is the decay parameter.

### i. Derivation of the change statistic $\delta_{GWD}(k, \alpha)$

The change statistic $\delta_{GWD}(k, \alpha)$ is the difference in the network statistic before and after adding a directed edge to a sender node that currently has out-degree $k$.

Let $x$ be the initial network and $x^+$ be the network after the edge is added. The addition of the edge causes $D_k^{out}$ to decrease by 1 and $D_{k+1}^{out}$ to increase by 1.

$$\delta_{GWD}(k, \alpha) = s_{GWD}(x^+, \alpha) - s_{GWD}(x, \alpha)$$
$$\delta_{GWD}(k, \alpha) = e^\alpha \left[ \left(1 - (1 - e^{-\alpha})^{k+1}\right) - \left(1 - (1 - e^{-\alpha})^k\right) \right]$$
$$\delta_{GWD}(k, \alpha) = e^\alpha \left[ 1 - (1 - e^{-\alpha})^{k+1} - 1 + (1 - e^{-\alpha})^k \right]$$
$$\delta_{GWD}(k, \alpha) = e^\alpha \left[ (1 - e^{-\alpha})^k - (1 - e^{-\alpha})^{k+1} \right]$$

Factoring out $(1 - e^{-\alpha})^k$:
$$\delta_{GWD}(k, \alpha) = e^\alpha (1 - e^{-\alpha})^k \left[ 1 - (1 - e^{-\alpha}) \right]$$
$$\delta_{GWD}(k, \alpha) = e^\alpha (1 - e^{-\alpha})^k (e^{-\alpha})$$
$$\mathbf{\delta_{GWD}(k, \alpha) = (1 - e^{-\alpha})^k}$$

### ii. Plotting the change statistic

```{r}
#| label: setup
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyr)

delta_gwd <- function(k, alpha) {
  (1 - exp(-alpha))^k
}

alphas <- c(-1, -0.5, 0, 0.1, 0.2, 0.3, 0.5, log(2), 1, 2, 3)
k_range <- 0:15

plot_data <- expand.grid(k = k_range, alpha = alphas) %>%
  mutate(delta = delta_gwd(k, alpha),
         alpha_label = factor(round(alpha, 2)))
```

```{r}
#| label: fig-negative
#| fig-cap: "Change statistics for negative decay values"
#| fig-height: 4
#| message: false
#| warning: false

p1 <- plot_data %>% filter(alpha == -1) %>%
  ggplot(aes(x = k, y = delta)) +
  geom_line(color = "red") + geom_point() +
  labs(title = expression(paste("Change Statistic for ", alpha, " = -1")),
       y = expression(delta[GWD]), x = "Current Out-degree k") +
  theme_minimal()

p2 <- plot_data %>% filter(alpha == -0.5) %>%
  ggplot(aes(x = k, y = delta)) +
  geom_line(color = "orange") + geom_point() +
  labs(title = expression(paste("Change Statistic for ", alpha, " = -0.5")),
       y = expression(delta[GWD]), x = "Current Out-degree k") +
  theme_minimal()

p1
p2
```

```{r}
#| label: fig-positive
#| fig-cap: "Change statistics for non-negative decay values"
#| fig-height: 5
#| message: false
#| warning: false

# Combined plot for non-negative alphas
plot_data %>% filter(alpha >= 0) %>%
  ggplot(aes(x = k, y = delta, color = alpha_label, group = alpha_label)) +
  geom_line(linewidth = 1) + 
  geom_point() +
  labs(title = "Change Statistic for Non-negative Alpha Values",
       y = expression(delta[GWD]), x = "Current Out-degree k",
       color = expression(alpha)) +
  theme_minimal()
```

### iii. Main takeaways from the figures

The figures illustrate how the decay parameter $\alpha$ fundamentally alters the "marginal contribution" of adding an edge to a node based on its existing out-degree. This relationship is governed by the derived change statistic $\delta_{GWD}(k, \alpha) = (1 - e^{-\alpha})^k$, whose behavior deserves careful examination across the entire parameter space, particularly through explicit asymptotic analysis.

**Asymptotic behavior for positive and zero values:** The limiting case $\lim_{\alpha \to +\infty} (1 - e^{-\alpha})^k = 1^k = 1$ reveals that as the decay parameter becomes very large, the change statistic converges to a constant 1 regardless of out-degree $k$. This means the gwodegree term effectively reduces to a simple edges count at infinity, treating all edges equivalently regardless of whether they come from a node with low or high out-degree. Conversely, at $\alpha = 0$, we have $\delta_{GWD}(k, 0) = (1-1)^k = 0^k$, which equals 1 for $k=0$ and 0 for $k>0$—a step function that only rewards whether a node has at least one outgoing edge. Between these extremes, for **non-negative alpha values ($\alpha \geq 0$)**, the change statistic exhibits "diminishing returns." When $0 < \alpha < \infty$, the base $(1 - e^{-\alpha})$ lies strictly between 0 and 1, creating a geometric decay as $k$ increases. The specific value of $\alpha$ controls the speed of this decay: a small $\alpha$ (e.g., 0.1) produces rapid decay where only the first few edges contribute meaningfully, while larger $\alpha$ (e.g., 3) produces slower decay and maintains substantial weight across a wider range of out-degrees.

**Problematic behavior for negative values:** In contrast, **negative alpha values ($\alpha < 0$)** lead to mathematical and practical instability. As $\alpha$ decreases below zero, the base $(1 - e^{-\alpha}) < 0$ becomes negative, causing the change statistic $(1 - e^{-\alpha})^k$ to oscillate between positive and negative values depending on whether $k$ is even or odd. Critically, $\lim_{\alpha \to -\infty} (1 - e^{-\alpha})^k$ does not converge; instead, the magnitude $|1 - e^{-\alpha}|^k \to \infty$ exponentially. For intermediate negative values like $\alpha = -1$, these oscillations already explode in magnitude. In the context of ERGMs, such behavior leads to **model degeneracy**, where the simulation fails to converge because the model's preferences become infinitely large or unstable, rendering negative $\alpha$ values unsuitable for modeling real-world social processes.

### iv. Capturing an activity effect

An "activity effect" in social networks describes a "rich-get-richer" or preferential attachment mechanism where actors who are already highly active (those with high out-degrees) are more likely to send additional edges. This creates concentration: a few hub nodes drive much of the network structure. To capture this effect specifically using the gwodegree framework requires understanding how the statistic's coefficient interacts with the decay parameter $\alpha$.

The fundamental insight is that the gwodegree statistic with positive $\alpha$ is inherently designed to model **diminishing returns** from the perspective of low-degree nodes. Recall that $\delta_{GWD}(k, \alpha) = (1 - e^{-\alpha})^k$ decays geometrically in $k$. When a positive coefficient is applied to this statistic, the model rewards networks where contributions are spread evenly—maximizing the utility of early, low-degree edges. To reverse this preference and instead reward networks where high-degree nodes dominate, the gwodegree coefficient must be **negative** (e.g., $\theta < 0$).

With a negative coefficient, the model penalizes high gwodegree values, effectively punishing the "spread-out" edge distribution and favoring its opposite: concentrated edges among a few active hubs. The choice of the specific positive $\alpha$ value determines the **scope and sensitivity** of this activity effect. A **small positive $\alpha$** (e.g., 0.1 to 0.3) creates a very sharp discriminatory effect, heavily penalizing networks with many low-degree nodes while strongly rewarding concentration. This focuses the activity incentive on the critical transition from inactivity to activity. A **larger $\alpha$** (e.g., 1.0 or higher) produces a smoother effect where the discrimination between low-degree and high-degree nodes is more gradual, allowing the activity preference to persist across a wider range of out-degrees. In practice, $\alpha$ is often fixed to a moderate value like $\ln(2) \approx 0.69$ to balance statistical identifiability with sufficient heterogeneity in the marginal contribution function. This choice enables stable parameter estimation while maintaining enough flexibility to capture meaningful activity-driven processes in empirical networks.



# (2) Effect of the gwodegree coefficient on the out-degree distribution.
### i. ERGM on Wave 1 edges
```{r}
#| message: false
#| warning: false
library(network)
library(ergm)

file_path <- "Assignment_02/Glasgow/f1.csv" 

adj_m1 <- as.matrix(read.csv(file_path, header = FALSE))

glasgow1 <- as.network(adj_m1, directed = TRUE)

m1 <- ergm(glasgow1 ~ edges)

summary(m1)
```

### ii. ERGM on Wave 1 gwodegree $\{ -2, 2\}$ with fixed $\alpha = 0.3$ 
```{r}
#| message: false
#| warning: false

m2_neg <- ergm(glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)),
               offset.coef = -2)

m2_pos <- ergm(glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)),
               offset.coef = 2)

summary(m2_neg)
summary(m2_pos)

```


### iii. Simulations
```{r}
#| label: fig-violin-plots
#| fig-cap: "Out-degree distributions across models"
#| message: false
#| warning: false

# Fitted coefficients for fixed-parameter simulations
coef_m1 <- coef(m1)
coef_m2_neg <- coef(m2_neg)
coef_m2_pos <- coef(m2_pos)

# Simulate
sim_m1 <- simulate_formula(glasgow1 ~ edges, coef = coef_m1, nsim = 100)
sim_m2_neg <- simulate_formula(glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)), 
                               coef = c(coef_m2_neg["edges"], 0), 
                               offset.coef = -2, nsim = 100)
sim_m2_pos <- simulate_formula(glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)), 
                               coef = c(coef_m2_pos["edges"], 0), 
                               offset.coef = 2, nsim = 100)

get_degree_df <- function(sim_list, model_name) {
  # Out-degrees for every node in every simulated network
  do.call(rbind, lapply(1:length(sim_list), function(i) {
    data.frame(
      degree = summary(sim_list[[i]] ~ odegree(0:15)),
      deg_val = 0:15,
      model = model_name,
      sim_id = i
    )
  }))
}

plot_df <- rbind(
  get_degree_df(sim_m1, "M1 (Edges Only)"),
  get_degree_df(sim_m2_neg, "M2 (Coef = -2)"),
  get_degree_df(sim_m2_pos, "M2 (Coef = 2)")
)

# Violin plot
ggplot(plot_df, aes(x = as.factor(deg_val), y = degree, fill = model)) +
  geom_violin(position = position_dodge(width = 0.8), alpha = 0.7) +
  labs(title = "Out-degree Distribution Comparison",
       x = "Out-degree",
       y = "Number of Nodes",
       fill = "Model Configuration") +
  theme_minimal()
```

### iv. Comparison of out-degree distributions across configurations

The distinction between the three model configurations—M1 (edges only), M2 with negative gwodegree coefficient (Coef = −2), and M2 with positive gwodegree coefficient (Coef = 2)—reveals how network-level structure emerges from local preferences encoded in exponential random graph model parameters. Understanding these differences requires comparing the observed out-degree distributions against what we would expect from a random network with identical density.

**M1 (Edges only)** serves as a baseline and represents a purely random graph model. The edges-only specification produces a Poisson-like out-degree distribution centered around the expected density: nodes have relatively homogeneous out-degrees with light tails, reflecting that edge placement is equiprobable and independent (beyond the global density constraint). This distribution is what we would expect from a random network where density alone determines structure.

**M2 with negative coefficient (Coef = −2)** suppresses the gwodegree statistic, actively penalizing networks with high values of the gwodegree term. Since gwodegree with positive $\alpha$ rewards spread-out, "ordinary" out-degree distributions, applying a negative coefficient inverts this preference. The resulting networks exhibit **increased concentration**: fewer nodes with very high out-degrees (hubs) and more nodes with very low out-degrees. This distribution deviates substantially from the random baseline, showing pronounced right skewness and centralization. Real-world social networks often exhibit such patterns where a small fraction of actors drive most communication.

**M2 with positive coefficient (Coef = 2)** amplifies gwodegree, rewarding networks where edges are more evenly distributed. The resulting out-degree distribution becomes **more uniform and dispersed**, with reduced concentration. Compared to M1, fewer hubs emerge and the distribution is flatter; compared to negative coefficient, the distributions are fundamentally opposite. This configuration would better capture scenarios where activity is democratized across many actors rather than concentrated.

The magnitude of these distributional shifts—relative to the random baseline (M1)—demonstrates how even modest parameters ($\theta = \pm 2$) substantially reshape emergent network topology. The gwodegree parameter thus provides a powerful mechanism for controlling the balance between centralization and decentralization in ERGM simulations.

