---
title: "Assignment 2: Geometrically Weighted Terms in ERGMs"
author: "-"
format: html
---

# Task 1: Geometrically Weighted Terms in ERGMs
## (1) Change statistic of gwodegree

The **gwodegree** effect is defined as:
$$s_{GWD}(x, \alpha) = e^\alpha \sum_{k=1}^{n} \{1 - (1 - e^{-\alpha})^k\} D_k^{out}(x)$$
where $D_k^{out}(x)$ is the number of nodes with exactly out-degree $k$ in network $x$, and $\alpha$ is the decay parameter.

### i. Derivation of the change statistic $\delta_{GWD}(k, \alpha)$

The change statistic $\delta_{GWD}(k, \alpha)$ is the difference in the network statistic before and after adding a directed edge to a sender node that currently has out-degree $k$.

Let $x$ be the initial network and $x^+$ be the network after the edge is added. The addition of the edge causes $D_k^{out}$ to decrease by 1 and $D_{k+1}^{out}$ to increase by 1.

$$\delta_{GWD}(k, \alpha) = s_{GWD}(x^+, \alpha) - s_{GWD}(x, \alpha)$$
$$\delta_{GWD}(k, \alpha) = e^\alpha \left[ \left(1 - (1 - e^{-\alpha})^{k+1}\right) - \left(1 - (1 - e^{-\alpha})^k\right) \right]$$
$$\delta_{GWD}(k, \alpha) = e^\alpha \left[ 1 - (1 - e^{-\alpha})^{k+1} - 1 + (1 - e^{-\alpha})^k \right]$$
$$\delta_{GWD}(k, \alpha) = e^\alpha \left[ (1 - e^{-\alpha})^k - (1 - e^{-\alpha})^{k+1} \right]$$

Factoring out $(1 - e^{-\alpha})^k$:
$$\delta_{GWD}(k, \alpha) = e^\alpha (1 - e^{-\alpha})^k \left[ 1 - (1 - e^{-\alpha}) \right]$$
$$\delta_{GWD}(k, \alpha) = e^\alpha (1 - e^{-\alpha})^k (e^{-\alpha})$$
$$\mathbf{\delta_{GWD}(k, \alpha) = (1 - e^{-\alpha})^k}$$

### ii. Plotting the change statistic

```{r}
#| label: setup
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyr)

delta_gwd <- function(k, alpha) {
  (1 - exp(-alpha))^k
}

alphas <- c(-1, -0.5, 0, 0.1, 0.2, 0.3, 0.5, log(2), 1, 2, 3)
k_range <- 0:15

plot_data <- expand.grid(k = k_range, alpha = alphas) %>%
  mutate(delta = delta_gwd(k, alpha),
         alpha_label = factor(round(alpha, 2)))
```

```{r}
#| label: fig-negative
#| fig-cap: "Change statistics for negative decay values"
#| fig-height: 4
#| message: false
#| warning: false

# Create the plots
p1 <- plot_data %>% filter(alpha == -1) %>%
  ggplot(aes(x = k, y = delta)) +
  geom_line(color = "red") + geom_point() +
  labs(title = expression(paste("Change Statistic for ", alpha, " = -1")),
       y = expression(delta[GWD]), x = "Current Out-degree k") +
  theme_minimal()

p2 <- plot_data %>% filter(alpha == -0.5) %>%
  ggplot(aes(x = k, y = delta)) +
  geom_line(color = "orange") + geom_point() +
  labs(title = expression(paste("Change Statistic for ", alpha, " = -0.5")),
       y = expression(delta[GWD]), x = "Current Out-degree k") +
  theme_minimal()

p1
p2
```

```{r}
#| label: fig-positive
#| fig-cap: "Change statistics for non-negative decay values"
#| fig-height: 5
#| message: false
#| warning: false

# Combined plot for non-negative alphas
plot_data %>% filter(alpha >= 0) %>%
  ggplot(aes(x = k, y = delta, color = alpha_label, group = alpha_label)) +
  geom_line(linewidth = 1) + 
  geom_point() +
  labs(title = "Change Statistic for Non-negative Alpha Values",
       y = expression(delta[GWD]), x = "Current Out-degree k",
       color = expression(alpha)) +
  theme_minimal()
```

### iii. Main takeaways from the figures

The figures illustrate how the decay parameter $\alpha$ fundamentally alters the "marginal contribution" of adding an edge to a node based on its existing out-degree. This relationship is governed by the derived change statistic $\delta_{GWD}(k, \alpha) = (1 - e^{-\alpha})^k$.

For **non-negative alpha values ($\alpha \geq 0$)**, the change statistic exhibits "diminishing returns." When $\alpha > 0$, the base $(1 - e^{-\alpha})$ is a fraction between 0 and 1. Consequently, as the current out-degree $k$ increases, the value of the change statistic decreases geometrically toward zero. This implies that adding the first edge to a node (where $k=0$) contributes significantly more to the total network statistic than adding an edge to a node that is already "active" (high $k$). 

The specific value of $\alpha$ controls the "speed" of this decay. A small $\alpha$ (e.g., 0.1) results in a very rapid decay, where only the first few edges contribute meaningfully to the statistic. As $\alpha$ increases (e.g., $\alpha = 3$), the decay becomes much slower, and the change statistic remains near 1 for a wider range of $k$. In the limit where $\alpha \to \infty$, the change statistic becomes a constant 1, effectively turning the `gwodegree` term into a simple `edges` count. At $\alpha = 0$, the statistic becomes a "step function" that only counts whether a node has at least one outgoing edge.

In contrast, **negative alpha values ($\alpha < 0$)** lead to mathematical instability. Because the base $(1 - e^{-\alpha})$ becomes negative, the change statistic oscillates between positive and negative values as $k$ increases. For $\alpha = -1$, the magnitude of these oscillations explodes exponentially. In the context of ERGMs, such behavior typically leads to **model degeneracy**, where the simulation fails to converge because the model's "preferences" become infinitely large or unstable, making it unsuitable for modeling real-world social processes.

### iv. Capturing an activity effect

An "activity" effect in social networks refers to a "rich-get-richer" process, where actors who are already active (those with high out-degrees) are more likely to send additional edges. To capture this specifically using the `gwodegree` framework, we must consider the relationship between the statistic $s_{GWD}$ and the model coefficient $\theta$.

As demonstrated in the plots, the `gwodegree` statistic with a positive $\alpha$ is designed to model **diminishing returns** (anti-centralization). It places higher weight on the formation of edges by low-degree nodes. Therefore, to capture an **activity effect** (centralization), the decay parameter $\alpha$ should be chosen as a **positive value**, but the model should be estimated with a **negative coefficient** ($\theta < 0$).

When the coefficient for `gwodegree` is negative, the model "penalizes" networks with high `gwodegree` values. Since a high `gwodegree` statistic represents a network where edges are spread out evenly among many nodes (maximizing the "utility" of low-degree edges), a negative coefficient rewards the opposite: a network where edges are concentrated in a few highly active hubs. 

The choice of the specific $\alpha$ value determines the "scope" of this activity effect. A **small positive $\alpha$** (e.g., 0.1 to 0.3) creates a very sharp distinction between low-degree and high-degree nodes, focusing the activity effect on the transition from being inactive to active. A **larger $\alpha$** (e.g., 1.0 or higher) smooths this effect, allowing the "activity" incentive to persist even as nodes reach higher degrees. In practice, $\alpha$ is often fixed to a value like $\ln(2) \approx 0.69$ to provide a balance that allows for stable estimation while still allowing high-degree nodes to drive the network structure through a negative coefficient.


# (2) Effect of the gwodegree coefficient on the out-degree distribution.
### i. ERGM on Wave 1 edges
```{r}
#| message: false
#| warning: false
library(network)
library(ergm)

file_path <- "Assignment_02/Glasgow/f1.csv" 

adj_m1 <- as.matrix(read.csv(file_path, header = FALSE))

glasgow1 <- as.network(adj_m1, directed = TRUE)

m1 <- ergm(glasgow1 ~ edges)

summary(m1)
```

### ii. ERGM on Wave 1 gwodegree $\{ -2, 2\}$ with fixed $\alpha = 0.3$ 
```{r}
#| message: false
#| warning: false

m2_neg <- ergm(glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)),
               offset.coef = -2)

m2_pos <- ergm(glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)),
               offset.coef = 2)

summary(m2_neg)
summary(m2_pos)

```


### iii. Simulations$ 
```{r}
#| label: fig-violin-plots
#| fig-cap: "Out-degree distributions across models"
#| message: false
#| warning: false

# 1. Simulate 100 networks from each model
sim_m1 <- simulate(m1, nsim = 100)
sim_m2_neg <- simulate(m2_neg, nsim = 100)
sim_m2_pos <- simulate(m2_pos, nsim = 100)

# 2. Function to get degree distributions from a list of simulated networks
get_degree_df <- function(sim_list, model_name) {
  # Get out-degrees for every node in every simulated network
  # This creates a long data frame for plotting
  do.call(rbind, lapply(1:length(sim_list), function(i) {
    data.frame(
      degree = summary(sim_list[[i]] ~ odegree(0:15)),
      deg_val = 0:15,
      model = model_name,
      sim_id = i
    )
  }))
}

# 3. Combine data for plotting
plot_df <- rbind(
  get_degree_df(sim_m1, "M1 (Edges Only)"),
  get_degree_df(sim_m2_neg, "M2 (Coef = -2)"),
  get_degree_df(sim_m2_pos, "M2 (Coef = 2)")
)

# 4. Create the violin plot
ggplot(plot_df, aes(x = as.factor(deg_val), y = degree, fill = model)) +
  geom_violin(position = position_dodge(width = 0.8), alpha = 0.7) +
  labs(title = "Out-degree Distribution Comparison",
       x = "Out-degree",
       y = "Number of Nodes",
       fill = "Model Configuration") +
  theme_minimal()
```