---
title: "Assignment 2: Geometrically Weighted Terms in ERGMs"
author: "-"
format: pdf
---

# Task 1: Geometrically Weighted Terms in ERGMs
## (1) Change statistic of gwodegree

The **gwodegree** effect is defined as:
$$s_{GWD}(x, \alpha) = e^\alpha \sum_{k=1}^{n} \{1 - (1 - e^{-\alpha})^k\} D_k^{out}(x)$$
where $D_k^{out}(x)$ is the number of nodes with exactly out-degree $k$ in network $x$, and $\alpha$ is the decay parameter.

### i. Derivation of the change statistic $\delta_{GWD}(k, \alpha)$

The change statistic $\delta_{GWD}(k, \alpha)$ is the difference in the network statistic before and after adding a directed edge to a sender node that currently has out-degree $k$.

Let $x$ be the initial network and $x^+$ be the network after the edge is added. The addition of the edge causes $D_k^{out}$ to decrease by 1 and $D_{k+1}^{out}$ to increase by 1.

$$\delta_{GWD}(k, \alpha) = s_{GWD}(x^+, \alpha) - s_{GWD}(x, \alpha)$$
$$\delta_{GWD}(k, \alpha) = e^\alpha \left[ \left(1 - (1 - e^{-\alpha})^{k+1}\right) - \left(1 - (1 - e^{-\alpha})^k\right) \right]$$
$$\delta_{GWD}(k, \alpha) = e^\alpha \left[ 1 - (1 - e^{-\alpha})^{k+1} - 1 + (1 - e^{-\alpha})^k \right]$$
$$\delta_{GWD}(k, \alpha) = e^\alpha \left[ (1 - e^{-\alpha})^k - (1 - e^{-\alpha})^{k+1} \right]$$

Factoring out $(1 - e^{-\alpha})^k$:
$$\delta_{GWD}(k, \alpha) = e^\alpha (1 - e^{-\alpha})^k \left[ 1 - (1 - e^{-\alpha}) \right]$$
$$\delta_{GWD}(k, \alpha) = e^\alpha (1 - e^{-\alpha})^k (e^{-\alpha})$$
$$\delta_{GWD}(k, \alpha) = (1 - e^{-\alpha})^k$$

### ii. Plotting the change statistic

```{r}
#| label: setup
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyr)

delta_gwd <- function(k, alpha) {
  (1 - exp(-alpha))^k
}

alphas <- c(-1, -0.5, 0, 0.1, 0.2, 0.3, 0.5, log(2), 1, 2, 3)
k_range <- 0:15

plot_data <- expand.grid(k = k_range, alpha = alphas) %>%
  mutate(delta = delta_gwd(k, alpha),
         alpha_label = factor(round(alpha, 2)))
```

```{r}
#| label: fig-negative
#| fig-cap: "Change statistics for negative decay values"
#| fig-height: 4
#| message: false
#| warning: false

p1 <- plot_data %>% filter(alpha == -1) %>%
  ggplot(aes(x = k, y = delta)) +
  geom_line(color = "red") + geom_point() +
  labs(title = expression(paste("Change Statistic for ", alpha, " = -1")),
       y = expression(delta[GWD]), x = "Current Out-degree k") +
  theme_minimal()

p2 <- plot_data %>% filter(alpha == -0.5) %>%
  ggplot(aes(x = k, y = delta)) +
  geom_line(color = "orange") + geom_point() +
  labs(title = expression(paste("Change Statistic for ", alpha, " = -0.5")),
       y = expression(delta[GWD]), x = "Current Out-degree k") +
  theme_minimal()

p1
p2
```

```{r}
#| label: fig-positive
#| fig-cap: "Change statistics for non-negative decay values"
#| fig-height: 5
#| message: false
#| warning: false

# Combined plot for non-negative alphas
plot_data %>% filter(alpha >= 0) %>%
  ggplot(aes(x = k, y = delta, color = alpha_label, group = alpha_label)) +
  geom_line(linewidth = 1) + 
  geom_point() +
  labs(title = "Change Statistic for Non-negative Alpha Values",
       y = expression(delta[GWD]), x = "Current Out-degree k",
       color = expression(alpha)) +
  theme_minimal()
```

### iii. Main takeaways from the figures

The figures illustrate how the decay parameter $\alpha$ fundamentally alters the "marginal contribution" of adding an edge to a node based on its existing out-degree. This relationship is governed by the derived change statistic $\delta_{GWD}(k, \alpha) = (1 - e^{-\alpha})^k$, whose behavior deserves careful examination across the entire parameter space, particularly through explicit asymptotic analysis.

**Asymptotic behavior for positive and zero values:** The limiting case $\lim_{\alpha \to +\infty} (1 - e^{-\alpha})^k = 1^k = 1$ reveals that as the decay parameter becomes very large, the change statistic converges to a constant 1 regardless of out-degree $k$. This means the gwodegree term effectively reduces to a simple edges count at infinity, treating all edges equivalently regardless of whether they come from a node with low or high out-degree. Conversely, at $\alpha = 0$, we have $\delta_{GWD}(k, 0) = (1-1)^k = 0^k$, which equals 1 for $k=0$ and 0 for $k>0$—a step function that only rewards whether a node has at least one outgoing edge. Between these extremes, for **non-negative alpha values ($\alpha \geq 0$)**, the change statistic exhibits "diminishing returns." When $0 < \alpha < \infty$, the base $(1 - e^{-\alpha})$ lies strictly between 0 and 1, creating a geometric decay as $k$ increases. The specific value of $\alpha$ controls the speed of this decay: a small $\alpha$ (e.g., 0.1) produces rapid decay where only the first few edges contribute meaningfully, while larger $\alpha$ (e.g., 3) produces slower decay and maintains substantial weight across a wider range of out-degrees.

**Problematic behavior for negative values:** In contrast, **negative alpha values ($\alpha < 0$)** lead to mathematical and practical instability. As $\alpha$ decreases below zero, the base $(1 - e^{-\alpha}) < 0$ becomes negative, causing the change statistic $(1 - e^{-\alpha})^k$ to oscillate between positive and negative values depending on whether $k$ is even or odd. Critically, $\lim_{\alpha \to -\infty} (1 - e^{-\alpha})^k$ does not converge; instead, the magnitude $|1 - e^{-\alpha}|^k \to \infty$ exponentially. For intermediate negative values like $\alpha = -1$, these oscillations already explode in magnitude. In the context of ERGMs, such behavior leads to **model degeneracy**, where the simulation fails to converge because the model's preferences become infinitely large or unstable, rendering negative $\alpha$ values unsuitable for modeling real-world social processes.

### iv. Capturing an activity effect

An "activity effect" in social networks describes a "rich-get-richer" or preferential attachment mechanism where actors who are already highly active (those with high out-degrees) are more likely to send additional edges. This creates concentration: a few hub nodes drive much of the network structure. To capture this effect specifically using the gwodegree framework requires understanding how the statistic's coefficient interacts with the decay parameter $\alpha$.

The fundamental insight is that the gwodegree statistic with positive $\alpha$ is inherently designed to model **diminishing returns** from the perspective of low-degree nodes. Recall that $\delta_{GWD}(k, \alpha) = (1 - e^{-\alpha})^k$ decays geometrically in $k$. When a positive coefficient is applied to this statistic, the model rewards networks where contributions are spread evenly—maximizing the utility of early, low-degree edges. To reverse this preference and instead reward networks where high-degree nodes dominate, the gwodegree coefficient must be **negative** (e.g., $\theta < 0$).

With a negative coefficient, the model penalizes high gwodegree values, effectively punishing the "spread-out" edge distribution and favoring its opposite: concentrated edges among a few active hubs. The choice of the specific positive $\alpha$ value determines the **scope and sensitivity** of this activity effect. A **small positive $\alpha$** (e.g., 0.1 to 0.3) creates a very sharp discriminatory effect, heavily penalizing networks with many low-degree nodes while strongly rewarding concentration. This focuses the activity incentive on the critical transition from inactivity to activity. A **larger $\alpha$** (e.g., 1.0 or higher) produces a smoother effect where the discrimination between low-degree and high-degree nodes is more gradual, allowing the activity preference to persist across a wider range of out-degrees. In practice, $\alpha$ is often fixed to a moderate value like $\ln(2) \approx 0.69$ to balance statistical identifiability with sufficient heterogeneity in the marginal contribution function. This choice enables stable parameter estimation while maintaining enough flexibility to capture meaningful activity-driven processes in empirical networks.



# (2) Effect of the gwodegree coefficient on the out-degree distribution.
### i. ERGM on Wave 1 edges
```{r}
#| message: false
#| warning: false
library(network)
library(ergm)

file_path <- "Assignment_02/Glasgow/f1.csv" 

adj_m1 <- as.matrix(read.csv(file_path, header = FALSE))

glasgow1 <- as.network(adj_m1, directed = TRUE)

m1 <- ergm(glasgow1 ~ edges)

summary(m1)
```

### ii. ERGM on Wave 1 gwodegree $\{ -2, 2\}$ with fixed $\alpha = 0.3$ 
```{r}
#| message: false
#| warning: false

m2_neg <- ergm(glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)),
               offset.coef = -2)

m2_pos <- ergm(glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)),
               offset.coef = 2)

summary(m2_neg)
summary(m2_pos)

```


### iii. Simulations
```{r}
#| label: fig-violin-plots
#| fig-cap: "Out-degree distributions across models"
#| fig-width: 10
#| fig-height: 5.5
#| message: false
#| warning: false

# Fitted coefficients for fixed-parameter simulations
coef_m1 <- coef(m1)
coef_m2_neg <- coef(m2_neg)
coef_m2_pos <- coef(m2_pos)

# Simulate
sim_m1 <- simulate_formula(glasgow1 ~ edges, coef = coef_m1, nsim = 100)

sim_m2_neg <- simulate_formula(
  glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)), 
  coef = c(coef_m2_neg["edges"], 0), 
  offset.coef = -2, nsim = 100)

sim_m2_pos <- simulate_formula(
  glasgow1 ~ edges + offset(gwodegree(decay = 0.3, fixed = TRUE)), 
  coef = c(coef_m2_pos["edges"], 0), 
  offset.coef = 2, nsim = 100)

get_degree_df <- function(sim_list, model_name) {
  # Extract actual out-degree of each node in each simulated network
  do.call(rbind, lapply(1:length(sim_list), function(i) {
    net <- sim_list[[i]]
    degrees <- rowSums(as.matrix(net))  # Out-degree for each node
    data.frame(
      degree = degrees,
      node = 1:length(degrees),
      model = model_name,
      sim_id = i
    )
  }))
}

plot_df <- rbind(
  get_degree_df(sim_m1, "M1 (Edges Only)"),
  get_degree_df(sim_m2_neg, "M2 (Coef = -2)"),
  get_degree_df(sim_m2_pos, "M2 (Coef = 2)")
)

# Calculate observed out-degrees from the actual network
obs_degrees <- rowSums(as.matrix(glasgow1))

# Compute summary statistics for each simulation
summary_df <- plot_df %>%
  group_by(sim_id, model) %>%
  summarise(mean_degree = mean(degree), 
            median_degree = median(degree),
            sd_degree = sd(degree),
            .groups = "drop")

# Create violin plot with boxplot overlay for clean visualization
ggplot(summary_df, aes(x = model, y = mean_degree, fill = model)) +
  geom_violin(alpha = 0.5, trim = FALSE, color = NA) +
  geom_boxplot(width = 0.2, alpha = 0.7, outlier.size = 1.5) +
  labs(title = "Out-degree Distribution Comparison Across Models",
       subtitle = "Violin shape shows distribution shape; boxplot shows quartiles from 100 simulations",
       x = "Model",
       y = "Mean Out-degree (per simulation)") +
  scale_fill_manual(values = c("M1 (Edges Only)" = "#FF7F50",
                               "M2 (Coef = -2)" = "#2ECC71",
                               "M2 (Coef = 2)" = "#3498DB")) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 0.7),
        plot.title = element_text(size = 16, face = "bold", margin = margin(b = 8)),
        plot.subtitle = element_text(size = 12, margin = margin(b = 12)),
        plot.margin = margin(t = 10, r = 20, b = 10, l = 10))
```

### iv. Comparison of out-degree distributions across configurations

The violin plot reveals striking empirical differences between the three model configurations, demonstrating how the gwodegree parameter's coefficient profoundly affects network topology.

**M1 (Edges Only)** establishes the baseline random graph model. With a mean out-degree around 3.5, this edges-only specification generates networks with relatively uncontrolled degree variation. The violin's intermediate width suggests moderate variation across simulations—this is what we obtain from pure density constraints without any degree-targeting mechanism.

**M2 with Coef = −2** produces dramatically different networks. The mean out-degree increases to approximately 4.2–4.3, substantially higher than M1. The tighter violin shape indicates more consistent networks across simulations. The negative coefficient, applied to an offset gwodegree term, creates a strong preference for networks with **higher average connectivity**. Nodes tend to have more outgoing edges on average, generating denser networks. This demonstrates that negative weights on gwodegree push toward greater network activity overall.

**M2 with Coef = 2** creates the opposite effect. Mean out-degree drops to approximately 3.0, notably lower than M1. The narrow, concentrated violin shape confirms high consistency across simulations. The positive coefficient on gwodegree creates networks with **lower average connectivity**, resulting in sparser structures. Despite being narrower than M1, this configuration generates qualitatively different networks—fewer edges are formed overall, and when they are, they are distributed with less variation.

**Key observation**: Both M2 configurations (regardless of sign) produce tighter distributions than M1, indicating that the gwodegree term constrains degree variation—networks become more "regular" or predictable. However, the **direction** and **magnitude** of coefficients determine whether this constraint manifests as hyperactivity (+4.2 mean degree) or hypoactivity (3.0 mean degree). This demonstrates the gwodegree parameter's utility in controlling both the average level of activity and the consistency of that activity across network realizations.

